---
title: "NYTimes Text Analysis"
sidebar: false
format:
  html: 
    toc: true
editor_options: 
  chunk_output_type: console
---
                
## Introduction
I've always had an interest in current events, and growing up, I read the newspaper every morning. As an International Relations concentrator, I continue to be interested in events occurring around the world. I decided to work with the New York Times dataset for this project because I am interested in how different international events and countries have been portrayed in the media in the past, and how the overall sentiment toward these countries and events has changed. I hope to explore how certain events, both domestically and internationally, have changed the sentiment or number of articles written on certain topics. The data for this project comes from the NYTimes dataset included in the RTextTools package. The dataset consists of the headlines of front-page NYTimes articles from 1996 to 2006. 

```{r}
#| include: FALSE

#load libraries
library(tidyverse)
library(RTextTools)
library(lubridate)
library(tidytext)
library(wordcloud)
library(wordcloud2)
library(reshape2)
library(scales)
library(widyr)

#load data and convert to tibble
data(NYTimes)
nytimes <- as_tibble(NYTimes)
```

## Common Words and Sentiments in NYTimes Articles

To begin my analysis, I started by making a wordcloud of the New York Times articles to see which topics continually reappeared in the headlines. 


```{r}
#tidying data 
tidy_times <- nytimes |>
  mutate(Subject = as.character(Subject),
         Date = as.character(Date),
         Title = as.character(Title)) |>
  unnest_tokens(word, Title, token = "words")

#creating data for wordcloud
times_wordcloud <- tidy_times |>
  anti_join(stop_words) |>
  count(word) |>
  filter(!str_detect(word, "\\d")) |>
  arrange(desc(n)) |>
  slice_head(n = 80)
  
#wordcloud
  wordcloud2(
  times_wordcloud, 
  size = 1, 
  shape = 'circle',
  minSize = 3,
  color = 'random-dark'
)

```
Based on this wordcloud, common topics that frequently appear in the headlines concern domestic politics, the economy, elections, and international conflicts and relations. I then decided to explore how the content of these words impacted the sentiment of articles. 

## Overall Article Sentiment By Week

To look at the overall sentiment of articles, I divided the articles into weekly sections, and then examined the overall sentiment of each of these weeks by decade.

```{r}
#loading sentiment data
bing_sentiments <- get_sentiments(lexicon = "bing")

#creating article by sentiment plot
tidy_times |>
  mutate(linenumber = row_number(),
         decade = ifelse(str_detect(Date, "^1.*"), "1990s", "2000s")) |>
  inner_join(bing_sentiments) |>
  count(decade, index = linenumber %/% 70, sentiment) |>
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) |>
  mutate(sentiment = positive - negative) |>
  ggplot(aes(x = index, y = sentiment, fill = decade)) +
    geom_col(show.legend = FALSE) +
    geom_hline(yintercept = 0, color = "black") +
    facet_wrap(~decade, ncol = 2, scales = "free_x") +
    theme_bw() +
    scale_fill_manual(values = c("1990s" = "#0047AB", "2000s" = "maroon")) +
    labs(
      title = "Overall Sentiment of NY Times Articles by Week",
      x = "Week",
      y = "Sentiment"
    )
```
Oftentimes, the context of these headlines are negative. When examining the overall sentiment of articles for every week from 1996 to 2006, the sentiment is overwhelming negative, with very few weeks across both decades displaying an overall positive sentiment. 

## Prevalence of Different IR Subjects Across Time

Considering my interest in International Relations, I decided to examine the prevalence of different countries in the news across time. For this analysis, I chose to focus on China, Iraq, and Russia, which are countries that are commonly discussed in International Relations and have been important to foreign affairs in recent decades. 

```{r}
#creating count of different subjects plot
nytimes |>
  mutate(Subject = str_to_lower(Subject),
         year = str_extract(as.character(Date), "\\d{2}$"),
         century = ifelse(str_detect(year, "^9."), "19", "20"),
         year = as.integer(str_c(century, year))) |>
  filter(str_detect(Subject, "(russia|china|iraq)")) |>
  mutate(ir_subject = str_extract(Subject, "(russia|china|iraq)"),
         ir_subject = str_to_title(ir_subject)) |>
  group_by(year, ir_subject) |>
  summarize(n = n(), .groups = "drop") |>
  ggplot(aes(x = year, y = n, fill = ir_subject)) +
    geom_col(position = "dodge", color = "black") +
    scale_y_continuous(breaks = pretty_breaks(n = 4)) +
    scale_fill_manual(values = c("China" = "maroon", "Iraq" = "gold", "Russia" = "#0047AB")) +
    labs(
      title = "Number of Occurences of International Relations Topics in NYTimes Articles",
      subtitle = "From 1996 to 2006",
      x = "Year",
      y = "Number of Appearances",
      fill = "IR Subject"
    )
```

Based on this visualization, the majority of stories have been written about Iraq during this time period, with a large spike in 2002 and 2003, correlating with the beginning of the Iraq war. More stories were written about Russia and China in the late 1990s compared to Iraq, but the number of these stories tapered off with the start of the Iraq war.

## Correlated Topics with IR Subjects

Expanding the number of states examined to also include Afghanistan, the Balkans, and the U.S., I wanted to look into the words most correlated with these countries in order to learn what topics were commonly written about when these countries were in the news. 

```{r}
nytimes$Title <- as.character(nytimes$Title)

#sectioning rows to article length
times_section_words <- nytimes |>
  select(Date, Title) |>
  mutate(section = row_number() %/% 1) |>
  unnest_tokens(word, Title) |>
  filter(!word %in% stop_words$word,
         !is.na(word))

#finding correlations of articles
word_cors <- times_section_words |>
  group_by(word) |>
  filter(n() >= 5) |>
  pairwise_cor(word, section, sort = TRUE)

#plot of words correlated with different subjects
word_cors |>
  filter(item1 %in% c("iraq", "china", "russia", "afghanistan", "balkans")) |>
  group_by(item1) |>
  slice_max(correlation, n = 6) |>
  ungroup() |>
  mutate(item2 = reorder(item2, correlation),
        item1 = str_to_title(item1)) |>
  ggplot(aes(item2, correlation)) +
    geom_bar(stat = "identity", fill = "maroon", color = "black") +
    facet_wrap(~item1, scales = "free") +
    coord_flip() +
    labs(title = "Words Most Commonly Correlated with International Relations Subjects",
         x = "Correlation",
         y = ""
         )
```

Based on these correlations, it appears that news stories about China and Russia were more likely to be about economic issues, which makes sense considering the growth of China's economy and Russia's economic transition following the collapse of the USSR occurring during this time period. On the other hand, those about Afghanistan, Iraq, and the Balkans referred to the conflicts occurring in these countries during this time period. Both the Afghanistan and Iraq wars began in the early 2000s, and the Balkans was recovering from the collapse of the former Yugoslavia in the 1990s. It is particularly interesting that the word "crisis" has such an extremely strong correlation (0.6) to the Balkans in comparison to all the other words. 

## Sentiments Associated with Different IR Subjects

Beyond just the words most correlated with these countries, I was interested in looking at the sentiment of the articles written about these countries. 

```{r}
#joining times data with bing sentiments
bing_word_counts <- tidy_times |>
  inner_join(get_sentiments("bing")) |>
  group_by(sentiment, Subject) |>
  count(word, sentiment, sort = TRUE) |>
  ungroup()

#creating plot of bing sentiments by subject
bing_word_counts |>
  filter(str_detect(Subject, "(russia|china|iraq|israel)")) |>
  mutate(ir_subject = str_extract(Subject, "(russia|china|iraq|israel)"),
         ir_subject = str_to_title(ir_subject)) |>
  group_by(sentiment) |>
  slice_max(n, n = 10) |>
  ggplot(aes(x = ir_subject, fill = sentiment)) +
    geom_bar(position = "fill", color = "black") +
    scale_fill_manual(values = c("positive" = "#0047AB", negative = "maroon")) +
    labs(title = "Sentiment of Articles Associated with IR Subjects",
         x = "IR Subject",
         y = "Proportion of Sentiment",
         fill = "Sentiment")

```
Based on this plot, the sentiment of articles concerning International Relations subjects are generally negative. This is particularly true of Russia and China, in which the proportion of negative articles written about these countries is 1. Surprisingly, Iraq had the most positive sentiment of associated articles, with almost half of all articles having a more positive sentiment, even though the Iraq war was happening at this time.

## Conclusion

Overall, the sentiment of New York Times Articles from 1996 to 2006 is generally quite negative. The articles about International Relations subjects continue this trend, with lots of negative sentiment associated with these articles, particularly articles written about Russia and China. In the future, further analysis of different countries or subjects could be interesting to examine the sentiment of these topics and how they have differed over time. Furthermore, looking into how the number and sentiment of articles written about Russia and China has changed would be fascinating considering the importance these countries hold in International Relations today. 
